Modal Abstractions for Virtualizing Memory Addresses

### 2024-07-01 examples.tex
+

\end{array}
\]
\vspace{-1em}
\caption{Global Address-Space Invariant in Figure \ref{fig:peraspaceinvariant} extended with a ghost map bookkeeping identity mappings }
  \label{fig:peraspaceinvariant_with_p2v_extension}
  \end{figure*}
Before Line 7 in Figure \ref{fig:calltopteinitialize}, to read the value referenced by physical address \textsf{entry}, first we open our the addresss-space invariant ($\mathcal{I}\textsf{ASpace}_{\textsf{id}}(\ptablestore,\Xi,m)$), and then obtain the virtual mapping out of the identity mapping $\Xi$ with the side condition
\begin{equation*}
 \ulcorner \textsf{qfrac} = 1 \leftrightarrow \; \lnot\textsf{entry\_present }(\vale) \urcorner \tag{*}
\end{equation*}
 assuring that looking up identity mappings are safe, because the full ownership asserts the non-presence of the entry which requires allocation a new physical page and setting the presence bit. By obtaining the identity mapping for \textsf{entry} in Figure \ref{fig:calltopteinitialize} -- ($\textsf{entry-KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{entry entry\_val}$ with page addr \textsf{entry}),
 \[
 \begin{array}{l}
\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi\setminus\{entry\},m) \ast \textsf{rbp-8} \mapsto_{\textsf{v}} \textsf{entry} \ast r8  \mapsto_{\textsf{r}} \textsf{entry} \textsf{\_} \ast rdi  \mapsto_{\textsf{r}} \textsf{entry} \ast \textsf{entry-KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{entry\_val} \ast \sumapaces\rtv\delta  \ast \sumpv\paddr\qfrac\true  }_{\rtv}
mov    [r8],rdi
\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m) \ast \textsf{rbp-8} \mapsto_{\textsf{v}} \textsf{entry} \ast r8  \mapsto_{\textsf{r}} \textsf{entry} \textsf{\_} \ast rdi  \mapsto_{\textsf{r}} \textsf{entry\_val} \ast  \textsf{entry-KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{entry\_val} \ast \sumapaces\rtv\delta  \ast \sumpv\paddr\qfrac\true  }_{\rtv}
\end{array}
 \]
 we are able to load the \textsf{entry\_val} into \textsf{rdi}, and check the presence bit via $and 0x1,rdi$.

@@

In case of absence, \textsf{pte\_initialize} is called for a physical-zeroed page (\textsf{kalloc} -- currently an axiom as it is not part of \textsf{pte} module but the allocator's) to be backing the physical frame address of the entry.
\mytodo{Discuss with Colin: v2p and spec on pteinitialize. After clarification expand this piece a little bit more. Right now we have physical pointsto returned from kalloc so pfn is directly set not through v2p}
Once we make sure that the frame of the page-table entry is backed with an allocated physical page, we can obtain the virtual address of the next level-page table (\textsf{next} of function \textsf{get\_next\_table} in Figure \ref{}) via looking up the \textit{identity mapping} from the physical address (the address value referenced by \textsf{entry->pfn}) to virtual address value to be returned by \textsf{get\_next\_table}.
\begin{figure}\footnotesize
\begin{lstlisting}[mathescape]
;;if (!entry->present) {
  ...

;;}
...
;;uintptr_t next_virt_addr = (uintptr_t) P2V(next_phys_addr);
  $\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m) \ast \ulcorner \textsf{OFF} + \paddr > 0 \urcorner \ast \texttt{rbx}\mapsto_{\textsf{r}} \paddr \ast \sumapaces\rtv\delta  \ast \sumpv\paddr\qfrac\true  }_{\rtv}$
add rbx, KERNBASE ;; shift physical address by identity mapping constant OFF
$\specline{\textsf{P} \ast \exists \qfraczero; \qfrac,\textsf{ w, val},  \texttt{rbx}\mapsto_{\textsf{r}} \paddr + \textsf{OFF} \ast \sumapaces\rtv\delta \ast \sumpv\paddr\qfrac\true \ast \vaddr \mapsto_{\texttt{vpte}} \{\textsf{q}, \textsf{qfrac}\} \paddr \;(\texttt{val})}$
$\specline{\ulcorner \vaddr = \paddr + \textsf{OFF} \urcorner \ast \ulcorner \textsf{qfrac} = 1 \leftrightarrow \; \lnot\textsf{entry\_present }(\vale) \urcorner \ast \sumwalkabs\vaddr\qfrac\paddr \ast (R \wand \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m)) }_{\rtv}$
\end{lstlisting}
\vspace{-1em}
\caption{Converting a physical address, whose exsitence is made certain, of a PTE to a virtual address (w/o instruction pointer or flag updates). $R$ abbreviates the resources
associated with \paddr\ by $\Xi$ in Figure \ref{fig:peraspaceinvariant_with_p2v_extension}.}
\label{fig:p2v}
\end{figure}

As part of the full page table walk verification, we have verified the critical step for a small x86-64 kernel, which is the physical-to-virtual
conversion (often appearing as a macro \texttt{P2V} inside the macro \textsf{ PTE\_PFN\_TO\_ADDR} in C source code). In our small kernel, as in larger kernels, \texttt{P2V} is actually just addition,
but the correctness of this simple instruction is quite subtle, as shown in Figure \ref{fig:p2v}.
Beyond a frame ($P$), the precondition requires the current address space invariant, a check that the addition will not overflow, the physical address to convert
(in \lstinline|rbx|), knowledge that the current address space is valid, and knowledge that the physical address is in a range dedicated
to page table entries (as this physical-to-virtual conversion is only guaranteed for page table memory). 
The latter uses an additional ghost set added to the per-address-space
invariant per Figure \ref{fig:peraspaceinvariant_with_p2v_extension}. The extended invariant owns, for each memory location dedicated to page tables, some fraction of
that physical memory, and also guarantees the offset is the correct calculation.
Subtly, the operation ensures the relevant table entry is readable, but the exact portion of ownership returned must be determined by inspecting the valid bit
of the value in memory --- so full ownership is returned only for unused entries.
This is a simple piece of code whose functionality is critical and whose correctness is highly non-trivial. No prior work engages with this problem.

>>>>

\end{array}
\]
\vspace{-1em}
\caption{Global Address-Space Invariant in Figure \ref{fig:peraspaceinvariant} extended with a ghost map bookkeeping identity mappings }
  \label{fig:peraspaceinvariant_with_p2v_extension}
\end{figure*}

As part of the full page table walk verification, we have verified the critical step for a small x86-64 kernel, which is the physical-to-virtual
conversion (often appearing as a macro \texttt{P2V} inside the macro \textsf{ PTE\_PFN\_TO\_ADDR} in C source code). In our small kernel, as in larger kernels, \texttt{P2V} is actually just addition,
but the correctness of this simple instruction is quite subtle, as shown in Figure \ref{fig:p2v} -- showing the section of the \textsf{get\_next\_table} in Figure \ref{calltopteinitialize} after the entry presence ensured (Lines 1-19 in Figure \ref{fig:calltopteinitialize}).
Beyond a frame, the precondition before Line 7 requires the current address space invariant, a check that the addition will not overflow, the physical address to convert
(in \lstinline|rax|), knowledge that the current address space is valid, and knowledge that the physical address is in a range dedicated
to page table entries (as this physical-to-virtual conversion is only guaranteed for page table memory). 
The latter uses an additional ghost set added to the per-address-space
invariant per Figure \ref{fig:peraspaceinvariant_with_p2v_extension}. The extended invariant owns, for each memory location dedicated to page tables, some fraction of
that physical memory, and also guarantees the offset is the correct calculation.
Subtly, the operation ensures the relevant table entry is readable, but the exact portion of ownership returned must be determined by inspecting the valid bit
of the value in memory --- so full ownership is returned only for unused entries.
This is a simple piece of code whose functionality is critical and whose correctness is highly non-trivial. No prior work engages with this problem.

Consequently, before Line 7 in Figure \ref{fig:calltopteinitialize}, to read the value referenced by physical address \textsf{entry}, first we open our the addresss-space invariant ($\mathcal{I}\textsf{ASpace}_{\textsf{id}}(\ptablestore,\Xi,m)$), and then obtain the virtual mapping out of the identity mapping $\Xi$ with the side condition
\begin{equation*}
 \ulcorner \textsf{qfrac} = 1 \leftrightarrow \; \lnot\textsf{entry\_present }(\vale) \urcorner \tag{*}
\end{equation*}
 assuring that looking up identity mappings are safe, because the full ownership asserts the non-presence of the entry which requires allocation a new physical page and setting the presence bit. By obtaining the identity mapping for \textsf{entry} in Figure \ref{fig:calltopteinitialize} -- ($\textsf{entry-KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{entry entry\_val}$ with page addr \textsf{entry}),
 \[
 \begin{array}{l}
\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi\setminus\{entry\},m) \ast \textsf{rbp-8} \mapsto_{\textsf{v}} \textsf{entry} \ast r8  \mapsto_{\textsf{r}} \textsf{entry} \textsf{\_} \ast rdi  \mapsto_{\textsf{r}} \textsf{entry} \ast \textsf{entry-KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{entry\_val} \ast \sumapaces\rtv\delta  \ast \sumpv\paddr\qfrac\true  }_{\rtv}
mov    [r8],rdi
\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m) \ast \textsf{rbp-8} \mapsto_{\textsf{v}} \textsf{entry} \ast r8  \mapsto_{\textsf{r}} \textsf{entry} \textsf{\_} \ast rdi  \mapsto_{\textsf{r}} \textsf{entry\_val} \ast  \textsf{entry-KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{entry\_val} \ast \sumapaces\rtv\delta  \ast \sumpv\paddr\qfrac\true  }_{\rtv}
\end{array}
 \]
 we are able to load the \textsf{entry\_val} into \textsf{rdi}, and check the presence bit via $and 0x1,rdi$.

@@

In case of absence, \textsf{pte\_initialize} is called for a physical-zeroed page (\textsf{kalloc} -- currently an axiom as it is not part of \textsf{pte} module but the allocator's) to be backing the physical frame address of the entry.
\mytodo{Discuss with Colin: v2p and spec on pteinitialize. After clarification expand this piece a little bit more. Right now we have physical pointsto returned from kalloc so pfn is directly set not through v2p}
Once we make sure that the frame of the page-table entry is backed with an allocated physical page, we can obtain the virtual address of the next level-page table (\textsf{next} of function \textsf{get\_next\_table} in Figure \ref{}) via looking up the \textit{identity mapping} from the physical address (the address value referenced by $\textsf{entry}\mapsto\textsf{pfn}$) to virtual address value to be returned by \textsf{get\_next\_table}.
\begin{figure}\footnotesize
\begin{lstlisting}[mathescape]
;;if (!entry->present) {
  ...

;;}
...
mov    rax,-0x18[rbp]
;;uintptr_t next_virt_addr = (uintptr_t) P2V(next_phys_addr);
mov    -0x18[rbp],rax
movabs $0x0,rcx            187: R_X86_64_64	KERN_BASE
add    rcx,rax
mov    rax,-0x20[rbp]
\vspace{-1em}
\caption{Converting a physical address, whose exsitence is made certain, of a PTE to a virtual address (w/o instruction pointer or flag updates). $R$ abbreviates the resources
associated with \paddr\ by $\Xi$ in Figure \ref{fig:peraspaceinvariant_with_p2v_extension}.}
\label{fig:p2v}
\end{figure}
